---
title: "Analyzing NYC Flight Data - HW2"
author: "Kinjal Ray"
output:
  html_document:
    code_folding: hide
---

```{r}
install.packages("nycflights13", repos = "http://cran.us.r-project.org")
library(nycflights13)
nyc <- nycflights13::flights
```

###1. 
```{r}
nyc$airgain <- nyc$arr_delay - nyc$dep_delay
```

###PART A - Do Airlines Gain Time?
```{r}
ifelse(mean(nyc$airgain, na.rm = TRUE) > 0, "NO, on average, airlines don't gain time in the air", "YES, on average, airlines do gain time")
```


###PART B - Average Airgain by Airline Carrier
```{r}
avg_airgain <- tapply(nyc$airgain, nyc$carrier, mean, na.rm = TRUE)
```

###PART C - Build bar chart
```{r}
avg_airgain <- sort(avg_airgain)
barplot(avg_airgain, main = "Average Airgain by Carrier", xlab = "Carrier", ylab = "Airgain")
```

###2. Lowest Departure Delay, Best Airgain
```{r}
avg_dep_delay <- tapply(nyc$dep_delay, nyc$origin, mean, na.rm = TRUE)
least_dep_delay <- names(avg_dep_delay[avg_dep_delay == min(avg_dep_delay)])

avg_airgain_bycarrier <- tapply(nyc$airgain, nyc$origin, mean, na.rm = TRUE)
best_airgain <- names(avg_airgain_bycarrier[avg_airgain_bycarrier == min(avg_airgain_bycarrier)])
```
`r least_dep_delay` has the least amount of departure delay on average.
`r best_airgain` has the best airgain.


###3. Best Orgination Airport based on Month of Travel
The best origination airport for the month of June is LaGuardia airport (LGA)
```{r}
library(dplyr)

Best_Origin <- function(month_of_travel) {
  
month_of_travel <- switch (month_of_travel, "January" = 1, "February" = 2, "March" = 3,"April"=  4, "May" = 5, "June" = 6, "July" = 7, "August" = 8, "September" = 9, "Ocotber" = 10, "November" = 11, "December" = 12)

  by_month <- filter(nyc, nyc$month == month_of_travel)
  by_origin <- tapply(by_month$arr_delay, by_month$origin, mean, na.rm = TRUE)
  best_choice <- names(by_origin[by_origin == min(by_origin)])
  return(best_choice)
}

Best_Origin("June")

```


###4. Best Time of Day for Travel
The best time of day for travel is morning. 
```{r}
atl_only <- filter(nyc, nyc$dest == "ATL")
atl_only$timeOfDay <- ifelse(atl_only$hour >= 4 & atl_only$hour < 12, "Morning",ifelse(atl_only$hour >= 12 & atl_only$hour < 16, "Afternoon", "Evening"))

least_arr_delay <- tapply(atl_only$arr_delay, atl_only$timeOfDay, mean, na.rm = TRUE)

names(least_arr_delay[least_arr_delay == min(least_arr_delay)])

```



###Relationship between Distance and Airgain 
Plot 1:Grouped by Carrier 
Plot 2:Grouped by Time of Departure
```{r}
library(ggplot2)
baseplot <- ggplot(nyc, aes(x = nyc$distance, y = nyc$airgain, color = nyc$carrier)) + labs(x = "Distance", y = "Airgain", colour = "Carriers")

baseplot + geom_point(na.rm = TRUE) + facet_wrap(~ nyc$origin)

nyc$timeOfDay <- ifelse(nyc$hour >= 4 & nyc$hour < 12, "Morning", ifelse(nyc$hour >= 12 & nyc$hour < 16, "Afternoon", "Evening"))

baseplot2 <- ggplot(nyc, aes(x = nyc$distance, y = nyc$airgain, colour = nyc$timeOfDay)) + labs(x = "Distance", y = "Airgain", colour = "Time of Day")

baseplot2 + geom_point(na.rm = TRUE) + facet_wrap(~ nyc$origin)

```



###Airgain by Carrier
This violin plot showcasing the distribution of flight air gain amounts grouped by carrier. Although most carriers hover around the 0 air gain amount, the plot provides useful insight on performance of carriers. For example, the OO airline seems more consistent than other airlines, because it has much shorter tails. YV, EV, B6, and 9E predominantly have negative air gain, meaning they finish flights faster than expected. 
```{r}

ggplot(nyc, aes(x = nyc$carrier, y = nyc$airgain, colour = nyc$carrier)) + geom_violin(na.rm = TRUE) + labs(x = "Carrier", y  = "Airgain", colour = "Carrier")

```


